MASCARA: coexpression analysis in data from designed experiments

### 2023-12-19 R/Paper_markdown.Rmd
23_11_12_2

Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text. Panel *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP). All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design. Note; URPs are not necessarily completely unlinked to POIs or RPs, in the biological sense, but are assumed here to be several degrees away. Far enough that the effect can be approximated orthogonally to the coexpression, when modelling. When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance. Between-group variance pertains to the differences between experimental groups (or conditions). This type of variance is of interest when conducting DE studies. Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE. It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype). Concretising this with a biological example; in @haiderTranscriptomeAnalysisPhosphate2023 a phosphate (P) starvation was applied to rice plants to induce strigolactone (SL) biosynthesis, to study the change in gene expression in comparison to control plants with normal supplies of P. The SL pathway is known to be triggered through P deficiency. However, as P is one of three key nutrients for plant growth, deficiency of this nutrient causes widespread differential expression throughout the transcriptome (@wangRoleStrigolactonesDeficiency2021).

@@

Pearson's correlation coefficient *r* is calculated between all columns of \(\bm{Y}\). Ranks are assigned to all non-bait transcripts for each bait based on *r*, with the lowest rank for the highest correlation. These ranks are averaged across all baits.

<!-- (see supplementary equation \@ref(eq:rankcor)). -->

@@

To evaluate performance we use log2 geometric mean rank (GMR; $\bar{\gamma}$), defined as:

@@

  ggtitle(methods[i]) +
  ylab("Effect Size") + 
  xlab("Num. DEs") +
  theme_bw() +
  theme(plot.title = element_text(size=8)) 
```

@@

```{r tc, echo = F, fig.cap= "Results of simulation 1 for all 4 methods. Log2 geometric mean rank (GMR) as a function of the number of differentially expressed URP genes (x-axis) and combined effect size (y axis); ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$), lower GMR; dark blue, indicates better performance. ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel)."}
wrap_plots(plots) + plot_layout(guides = "collect")

@@

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method)) +
  scale_color_viridis(discrete = T)+
    stat_summary() +
    ylim(0,max ) +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  theme_bw() +
      theme(axis.title = element_blank())

@@

# ps1 <- grid.arrange(ps, left = "Log2(GMR)", bottom = "Replicate Number                     ")

a <- knitr::include_graphics("C:/Users/fwhite/Documents/replication_res.png")
a

@@

<!-- Ranked correlations are calculated from the correlation matrix $\mathbf{C} as: -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- RANKS  -->
<!-- \end{aligned}(\#eq:rankcor) -->
<!-- \end{equation} -->

To calculate Fisher transformed Pearson correlations (z) with Pearson's r:

@@

<!--   r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}} -->

>>>>

Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text. Panel *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP). All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design. URPs defined to be pathways not directly interacting with the POI. When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance. Between-group variance pertains to the differences between experimental groups (or conditions). This type of variance is of interest when conducting DE studies. Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE. It is the excessive between-group variance that leads to the densely connected nature of panel *B* from \@ref(fig:nettype). Concretising this with a biological example; in @haiderTranscriptomeAnalysisPhosphate2023 a phosphate (P) starvation was applied to rice plants to induce strigolactone (SL) biosynthesis, to study the change in gene expression in comparison to control plants with normal supplies of P. The SL pathway is known to be triggered through P deficiency. However, as P is one of three key nutrients for plant growth, deficiency of this nutrient causes widespread differential expression throughout the transcriptome (@wangRoleStrigolactonesDeficiency2021).

@@

Pearson's correlation coefficient *r* is calculated between all columns of \(\bm{Y}\). Ranks are assigned to all non-bait transcripts for each bait based on *r*, with the lowest rank for the highest correlation. These ranks are averaged across all baits (see supplementary equation \@ref(eq:rankcor)).

@@

To evaluate performance we use log2 geometric mean rank (log2(GMR); $\bar{\gamma}$), defined as:

@@

  ggtitle(methods[i]) +
  ylab("Effect Size") + 
  xlab("Num. DEs") +
  theme_bw() +
  labs(fill = "log2(GMR)") +
  theme(plot.title = element_text(size=8)) 
```

@@

```{r tc, echo = F, fig.cap= "Results of simulation 1 for all 4 methods. Log2 geometric mean rank (log2(GMR)) as a function of the number of differentially expressed URP genes (x-axis) and combined effect size (y axis); ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$), lower log2(GMR); dark blue, indicates better performance. ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel)."}
wrap_plots(plots) + plot_layout(guides = "collect")

@@

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method)) +
  scale_color_viridis(discrete = T)+
    stat_summary() +
    ylim(0,max ) +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  theme_bw() +
      theme(axis.title = element_blank())

@@

ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")

a <- knitr::include_graphics("replication_res.png")
a

@@

## Ranked correlations
Ranked correlations are calculated as:

\begin{equation}
\begin{aligned}
Rank_{\tilde{J}} = \frac{\sum_{b=1}^{B}rank_{b\tilde{j}}}{B}  
\end{aligned}(\#eq:rankcor)
\end{equation}

Where *B* is the number of baits and $rank_{b\tilde{j}}$ is the ranked Pearson correlation coefficient between gene j and bait b in the context of $\tilde{J}$ non-bait genes.

## Fisher transformed Pearson correlation coefficient
To calculate Fisher transformed Pearson correlations (z) with Pearson's r:

@@

<!-- $r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}$ -->
